Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines

### 2025-07-03 docs/paper.Rmd
write q-q evidence in the philippines and southeast asia

Empirical investigations of the quantity–quality trade-off have focused primarily on educational outcomes. Studies in both high-income and low-income settings have examined the effects of fertility on school enrollment, grade progression, test scores, and completed years of schooling \citep{rosenzweig1980testing, black2005more, angrist2010effects}. These outcomes serve as accessible proxies for long-run human capital accumulation, but they represent only one dimension of child quality. Other outcomes, such as nutritional status, are early-onset or biologically constrained. They also tend to be less responsive to remediation later in life. A child who suffers from chronic malnutrition may exhibit permanently reduced cognitive capacity and face limits in physical development that affect long-run productivity regardless of subsequent educational access \citep{hoddinott2013adult, grantham2007development}.

@@

\subsubsection{Recent Refinements to the Q-Q Model}

@@

>>>>

Empirical investigations of the quantity–quality trade-off have focused primarily on educational outcomes. Studies in both high-income and low-income settings have examined the effects of fertility on school enrollment, grade progression, test scores, and completed years of schooling \citep{rosenzweig1980testing, black2005more, angrist2010multiple}. These outcomes serve as accessible proxies for long-run human capital accumulation, but they represent only one dimension of child quality. Other outcomes, such as nutritional status, are early-onset or biologically constrained. They also tend to be less responsive to remediation later in life. A child who suffers from chronic malnutrition may exhibit permanently reduced cognitive capacity and face limits in physical development that affect long-run productivity regardless of subsequent educational access \citep{hoddinott2013adult, grantham2007development}.

@@

\subsection{Recent Refinements to the Q-Q Model}

@@

\subsection{Philippine Evidence on the Q–Q Trade-off}

Early empirical work in the Philippines provided suggestive support for this theory. For example, an influential study by \citet{horton1986child} used Philippine household data to jointly examine fertility and child nutrition and treated nutritional status as a measure of child quality. Horton found that better-educated mothers and fathers tended to choose smaller families and achieved better-nourished children–evidence of substitution away from “quantity” toward child “quality.” Notably, she observed strong birth-order effects on nutrition (first-born children faring better than later-born), which hint that parents may not evenly distribute quality investments among all children. This early work indicated a quantity–quality (Q–Q) trade-off in Philippine families, though it largely documented correlations rather than definitive causation.

Subsequent studies in the Philippines have tackled the critical issue of causality, using innovative research designs to isolate exogenous changes in family size. One seminal contribution is \citet{orbeta2010number}, who examined the impact of family size on children’s schooling using a nationally representative survey. Recognizing that family size is endogenous (parents’ fertility choices may reflect unobserved preferences or constraints), Orbeta employed an instrumental-variable (IV) approach grounded in Becker’s framework. Specifically, he used the sex composition of the first two children as an instrument for having additional children – an approach pioneered by \citet{angrist1996children} – leveraging the fact that Filipino parents often desire a mixed-gender sibset (and are more likely to have a third child if the first two are the same sex) \citep{vicerra2013fertility}. This strategy aims to generate random-like variation in family size uncorrelated with parental characteristics.

The IV estimates confirmed a significant negative causal effect of higher fertility on educational outcomes. \citet{orbeta2010number} found that each additional child in the household reduced the proportion of school-age children (6–24 years) attending school by roughly 19% of the baseline attendance rate. The trade-off was especially pronounced at higher education levels: for example, the estimated drop in school attendance was about 26% at the secondary level and 57% at the tertiary level for each additional sibling. These are sizable effects, implying that children from large families are substantially less likely to remain in school, presumably due to tighter household budget constraints or diluted parental attention. Moreover, the burden of the trade-off appeared regressive: Orbeta’s results showed much larger schooling deficits from an extra child in poorer households than in richer ones. For instance, in the poorest quintile, an additional sibling reduced school attendance by an estimated 24% (for ages 6–24), compared to a 16% reduction in the richest quintile. This regressive pattern aligns with Becker’s theory that resource constraints bind more tightly for low-income families, which makes the Q–Q trade-off more acute. In summary, Orbeta’s study – the first in the Philippines to account for fertility endogeneity – provides robust evidence that increases in family size cause significant declines in child educational attainment and validates the quantity–quality trade-off in this context. Large family size thus emerges as one mechanism contributing to poverty, by impeding children’s human capital accumulation in the Philippines.

Further compelling evidence comes from a natural experiment studied by \citet{dumas2019sex}. They exploit a unique policy shock in metropolitan Manila to identify the trade-off. In 1998, the Mayor of Manila city imposed a sudden ban on modern contraceptives in public facilities, drastically curtailing access to family planning for residents of Manila city (but not in surrounding municipalities). Dumas and Lefranc use this policy as a quasi-experiment: comparing families in Manila city (treated by the ban) to similar families in other cities unaffected by the ban before and after 1998. This difference-in-differences design, coupled with the fact that older mothers were naturally less fecund during the ban, isolates an exogenous fertility increase.

The results are striking. The contraceptives ban led to a significant rise in births and family size in Manila city relative to the control areas. Correspondingly, children born in Manila during the ban era experienced a sizable decline in educational attainment – clear evidence of a Q–Q trade-off precipitated by the shock. In the authors’ words, the policy-driven increase in family size “provide[s] evidence of a quality–quantity trade-off”: larger families, forced by the ban, resulted in lower schooling outcomes per child. By leveraging an actual anti-contraception policy, this study offers causal confirmation that constraints on family planning can worsen child quality. Together with Orbeta’s findings, it reinforces the relevance of Becker’s hypothesis in the Philippine setting: whether via unintended fertility (Manila’s case) or ordinary variation in sibling sex mix, more children mean fewer resources per child and thus worse education outcomes. Consistent with theory, Filipino parents facing exogenous increases in quantity were unable to maintain the same level of quality per child.

Beyond education, researchers have also examined health and other child outcomes in relation to family size. Evidence generally suggests the Q–Q trade-off extends to child health and nutrition. For instance, \citet{horton1986child} already hinted that large families may compromise child nutrition for later-born siblings. More recent regional research resonates with this. \citet{hatton2018fertility} analyze the effect of fertility on child height (a long-run health indicator) using longitudinal data from Indonesia – a neighboring Southeast Asian country with comparable developmental challenges. They address endogeneity by exploiting Indonesia’s family planning program rollout and exposure to mass media as instruments for fertility. The authors find a significant negative impact of family size on child health: each additional sibling is associated with about a one-third standard deviation reduction in a child’s height-for-age Z-score, after controlling for other factors. This health penalty from having more children is strongest in low-education households and appears in both urban and rural areas. Such findings mirror the Philippine evidence that the harms of large family size are most pronounced among disadvantaged families. In economic terms, poorer parents with many children struggle to provide adequate nutrition and schooling to all, which highlight the equity dimension of the Q–Q trade-off.

\subsection{Family Size and Child Outcomes in Southeast Asia}

The Philippine experience is echoed in other ASEAN countries, where researchers have probed the quantity–quality trade-off with diverse outcomes and methods. In Vietnam, for example, rapid fertility decline alongside rising education led to questions about a Q–Q mechanism. \citet{anh1998family} documented a negative correlation between family size and children’s school enrollment in Vietnam, though their analysis could not fully establish causality.

More rigorously, \citet{dang2016decision} used distance to the nearest family planning center as an instrument to study Vietnamese households’ investments in education. They introduced a novel measure of child quality—spending on private tutoring, a prevalent form of educational investment in Vietnam—alongside traditional indicators like schooling expenses. The IV estimates confirmed that children with more siblings receive significantly lower educational investments from their families. In particular, Vietnamese families of larger size spent less on each school-age child’s schooling and tutoring, even after controlling for community factors. This effect was robust across different definitions of family size and model specifications, indicating that Vietnamese parents do indeed trade off quantity for quality when faced with resource constraints. Such evidence aligns squarely with Becker’s model: as Vietnamese family size increases, per-child education spending falls; parents are prioritizing “quality” less when they have more offspring.

Indonesia and Thailand show similar patterns. Demographic research in Thailand during its fertility transition found that large families had markedly worse educational outcomes. \citet{knodel1990family}, studying Thai data in the 1990s, observed that once family size exceeded about 4–5 children, the likelihood of a child progressing to or staying in secondary school dropped precipitously compared to smaller families. Although these early Thai studies were based on correlations, they strongly suggested that limited family resources were being spread thin in big families and hurt children’s schooling attainment.

In Indonesia, cohort analyses and natural experiments reinforce the trade-off. \citet{maralani2008changing} showed that the relationship between sibship size and schooling evolved from neutral or even positive for older cohorts (born when education opportunities were limited) to negative for more recent cohorts, consistent with a growing importance of education in a modernizing economy. More concretely, the aforementioned study by \citet{hatton2018fertility} in Indonesia provides causal evidence that mirrors the Philippine findings in health and nutrition. Likewise, an analysis of Indonesian census data \citep{feng2021effect} found that having additional siblings significantly lowers children’s educational attainment once birth order effects are accounted for, paralleling results from China and Vietnam.

These regional studies share a commonality: in resource-constrained settings of Southeast Asia, increased child quantity tends to come at the expense of child quality, be it years of schooling, academic spending, or health status. The consistency of this pattern—across countries with different cultures and policies—highlights the fundamental economic logic identified by \citet{becker1973interaction}. Parents with finite resources face difficult choices, and many appear to balance quantity and quality in a way that confirms the trade-off hypothesis.

Despite these similarities, there are some noteworthy nuances and gaps in the ASEAN literature. One is the role of public policy and development level. Evidence suggests that the Q–Q trade-off may be mitigated in contexts with strong public support for education and health. For instance, studies in developed countries (e.g. Israel, Norway) often find little or no trade-off once factors like birth order are accounted for \citep{black2005more, kristensen2010educational, angrist2010multiple}. In Southeast Asia, however, public education quality and social safety nets are still developing, and the cost of raising children (education fees, food, etc.) is largely borne by families themselves \citep{oecd2024sigi}. This may explain why the trade-off emerges so clearly in the Philippines, Vietnam, Indonesia, and Thailand.

Another nuance is methodological: more recent studies employ credible identification strategies (IVs, twins, policy shocks) and consistently find a causal negative effect of family size on child outcomes, whereas older studies without such controls sometimes found weaker effects or none at all. This highlights the importance of accounting for endogeneity.

Finally, there remain gaps for future research. Most ASEAN studies focus on education and early-life health indicators as measures of child quality; there is relatively little evidence on long-term outcomes such as children’s eventual earnings or income in adulthood. It is not yet fully clear whether the schooling and health disadvantages observed in larger families translate into significantly lower adult productivity or income—a link that \citet{becker1976child} theorized but which could be explored further in this region.

Additionally, while the trade-off appears pervasive, its magnitude can vary: for example, the penalty of an extra child may be larger in poorer rural areas than in urban or wealthier settings, suggesting that local context (poverty, gender norms, access to services) can modulate the trade-off. Comparative studies across ASEAN are still somewhat limited, and a common challenge is disentangling related factors like birth order, sibling composition, and parental preferences.

Nonetheless, the prevailing evidence from the Philippines and its regional neighbors strongly supports Becker’s quantity–quality conjecture. As families have fewer children, they appear to invest more in each child’s education—investments crucial for human capital development and economic growth. Conversely, high-fertility households risk under-investing per child, which reinforces cycles of poverty and inequality. This literature thus provides an important empirical foundation for policies in the Philippines and ASEAN—from family planning programs to education subsidies—that aim to ease the quantity–quality trade-off and help families achieve both manageable size and better outcomes for the next generation.

### 2025-07-03 docs/paper.tex
write q-q evidence in the philippines and southeast asia

Empirical investigations of the quantity--quality trade-off have focused
primarily on educational outcomes. Studies in both high-income and
low-income settings have examined the effects of fertility on school
enrollment, grade progression, test scores, and completed years of
schooling
\citep{rosenzweig1980testing, black2005more, angrist2010effects}. These
outcomes serve as accessible proxies for long-run human capital
accumulation, but they represent only one dimension of child quality.
Other outcomes, such as nutritional status, are early-onset or
biologically constrained. They also tend to be less responsive to
remediation later in life. A child who suffers from chronic malnutrition
may exhibit permanently reduced cognitive capacity and face limits in
physical development that affect long-run productivity regardless of
subsequent educational access
\citep{hoddinott2013adult, grantham2007development}.

@@

>>>>

Empirical investigations of the quantity--quality trade-off have focused
primarily on educational outcomes. Studies in both high-income and
low-income settings have examined the effects of fertility on school
enrollment, grade progression, test scores, and completed years of
schooling
\citep{rosenzweig1980testing, black2005more, angrist2010multiple}. These
outcomes serve as accessible proxies for long-run human capital
accumulation, but they represent only one dimension of child quality.
Other outcomes, such as nutritional status, are early-onset or
biologically constrained. They also tend to be less responsive to
remediation later in life. A child who suffers from chronic malnutrition
may exhibit permanently reduced cognitive capacity and face limits in
physical development that affect long-run productivity regardless of
subsequent educational access
\citep{hoddinott2013adult, grantham2007development}.

@@

\subsection{Philippine Evidence on the Q–Q Trade-off}

Early empirical work in the Philippines provided suggestive support for
this theory. For example, an influential study by
\citet{horton1986child} used Philippine household data to jointly
examine fertility and child nutrition and treated nutritional status as
a measure of child quality. Horton found that better-educated mothers
and fathers tended to choose smaller families and achieved
better-nourished children--evidence of substitution away from
``quantity'' toward child ``quality.'' Notably, she observed strong
birth-order effects on nutrition (first-born children faring better than
later-born), which hint that parents may not evenly distribute quality
investments among all children. This early work indicated a
quantity--quality (Q--Q) trade-off in Philippine families, though it
largely documented correlations rather than definitive causation.

Subsequent studies in the Philippines have tackled the critical issue of
causality, using innovative research designs to isolate exogenous
changes in family size. One seminal contribution is
\citet{orbeta2010number}, who examined the impact of family size on
children's schooling using a nationally representative survey.
Recognizing that family size is endogenous (parents' fertility choices
may reflect unobserved preferences or constraints), Orbeta employed an
instrumental-variable (IV) approach grounded in Becker's framework.
Specifically, he used the sex composition of the first two children as
an instrument for having additional children -- an approach pioneered by
\citet{angrist1996children} -- leveraging the fact that Filipino parents
often desire a mixed-gender sibset (and are more likely to have a third
child if the first two are the same sex) \citep{vicerra2013fertility}.
This strategy aims to generate random-like variation in family size
uncorrelated with parental characteristics.

The IV estimates confirmed a significant negative causal effect of
higher fertility on educational outcomes. \citet{orbeta2010number} found
that each additional child in the household reduced the proportion of
school-age children (6--24 years) attending school by roughly 19\% of
the baseline attendance rate. The trade-off was especially pronounced at
higher education levels: for example, the estimated drop in school
attendance was about 26\% at the secondary level and 57\% at the
tertiary level for each additional sibling. These are sizable effects,
implying that children from large families are substantially less likely
to remain in school, presumably due to tighter household budget
constraints or diluted parental attention. Moreover, the burden of the
trade-off appeared regressive: Orbeta's results showed much larger
schooling deficits from an extra child in poorer households than in
richer ones. For instance, in the poorest quintile, an additional
sibling reduced school attendance by an estimated 24\% (for ages 6--24),
compared to a 16\% reduction in the richest quintile. This regressive
pattern aligns with Becker's theory that resource constraints bind more
tightly for low-income families, which makes the Q--Q trade-off more
acute. In summary, Orbeta's study -- the first in the Philippines to
account for fertility endogeneity -- provides robust evidence that
increases in family size cause significant declines in child educational
attainment and validates the quantity--quality trade-off in this
context. Large family size thus emerges as one mechanism contributing to
poverty, by impeding children's human capital accumulation in the
Philippines.

Further compelling evidence comes from a natural experiment studied by
\citet{dumas2019sex}. They exploit a unique policy shock in metropolitan
Manila to identify the trade-off. In 1998, the Mayor of Manila city
imposed a sudden ban on modern contraceptives in public facilities,
drastically curtailing access to family planning for residents of Manila
city (but not in surrounding municipalities). Dumas and Lefranc use this
policy as a quasi-experiment: comparing families in Manila city (treated
by the ban) to similar families in other cities unaffected by the ban
before and after 1998. This difference-in-differences design, coupled
with the fact that older mothers were naturally less fecund during the
ban, isolates an exogenous fertility increase.

The results are striking. The contraceptives ban led to a significant
rise in births and family size in Manila city relative to the control
areas. Correspondingly, children born in Manila during the ban era
experienced a sizable decline in educational attainment -- clear
evidence of a Q--Q trade-off precipitated by the shock. In the authors'
words, the policy-driven increase in family size ``provide{[}s{]}
evidence of a quality--quantity trade-off'': larger families, forced by
the ban, resulted in lower schooling outcomes per child. By leveraging
an actual anti-contraception policy, this study offers causal
confirmation that constraints on family planning can worsen child
quality. Together with Orbeta's findings, it reinforces the relevance of
Becker's hypothesis in the Philippine setting: whether via unintended
fertility (Manila's case) or ordinary variation in sibling sex mix, more
children mean fewer resources per child and thus worse education
outcomes. Consistent with theory, Filipino parents facing exogenous
increases in quantity were unable to maintain the same level of quality
per child.

Beyond education, researchers have also examined health and other child
outcomes in relation to family size. Evidence generally suggests the
Q--Q trade-off extends to child health and nutrition. For instance,
\citet{horton1986child} already hinted that large families may
compromise child nutrition for later-born siblings. More recent regional
research resonates with this. \citet{hatton2018fertility} analyze the
effect of fertility on child height (a long-run health indicator) using
longitudinal data from Indonesia -- a neighboring Southeast Asian
country with comparable developmental challenges. They address
endogeneity by exploiting Indonesia's family planning program rollout
and exposure to mass media as instruments for fertility. The authors
find a significant negative impact of family size on child health: each
additional sibling is associated with about a one-third standard
deviation reduction in a child's height-for-age Z-score, after
controlling for other factors. This health penalty from having more
children is strongest in low-education households and appears in both
urban and rural areas. Such findings mirror the Philippine evidence that
the harms of large family size are most pronounced among disadvantaged
families. In economic terms, poorer parents with many children struggle
to provide adequate nutrition and schooling to all, which highlight the
equity dimension of the Q--Q trade-off.

\subsection{Family Size and Child Outcomes in Southeast Asia}

The Philippine experience is echoed in other ASEAN countries, where
researchers have probed the quantity--quality trade-off with diverse
outcomes and methods. In Vietnam, for example, rapid fertility decline
alongside rising education led to questions about a Q--Q mechanism.
\citet{anh1998family} documented a negative correlation between family
size and children's school enrollment in Vietnam, though their analysis
could not fully establish causality.

More rigorously, \citet{dang2016decision} used distance to the nearest
family planning center as an instrument to study Vietnamese households'
investments in education. They introduced a novel measure of child
quality---spending on private tutoring, a prevalent form of educational
investment in Vietnam---alongside traditional indicators like schooling
expenses. The IV estimates confirmed that children with more siblings
receive significantly lower educational investments from their families.
In particular, Vietnamese families of larger size spent less on each
school-age child's schooling and tutoring, even after controlling for
community factors. This effect was robust across different definitions
of family size and model specifications, indicating that Vietnamese
parents do indeed trade off quantity for quality when faced with
resource constraints. Such evidence aligns squarely with Becker's model:
as Vietnamese family size increases, per-child education spending falls;
parents are prioritizing ``quality'' less when they have more offspring.

Indonesia and Thailand show similar patterns. Demographic research in
Thailand during its fertility transition found that large families had
markedly worse educational outcomes. \citet{knodel1990family}, studying
Thai data in the 1990s, observed that once family size exceeded about
4--5 children, the likelihood of a child progressing to or staying in
secondary school dropped precipitously compared to smaller families.
Although these early Thai studies were based on correlations, they
strongly suggested that limited family resources were being spread thin
in big families and hurt children's schooling attainment.

In Indonesia, cohort analyses and natural experiments reinforce the
trade-off. \citet{maralani2008changing} showed that the relationship
between sibship size and schooling evolved from neutral or even positive
for older cohorts (born when education opportunities were limited) to
negative for more recent cohorts, consistent with a growing importance
of education in a modernizing economy. More concretely, the
aforementioned study by \citet{hatton2018fertility} in Indonesia
provides causal evidence that mirrors the Philippine findings in health
and nutrition. Likewise, an analysis of Indonesian census data
\citep{feng2021effect} found that having additional siblings
significantly lowers children's educational attainment once birth order
effects are accounted for, paralleling results from China and Vietnam.

These regional studies share a commonality: in resource-constrained
settings of Southeast Asia, increased child quantity tends to come at
the expense of child quality, be it years of schooling, academic
spending, or health status. The consistency of this pattern---across
countries with different cultures and policies---highlights the
fundamental economic logic identified by \citet{becker1973interaction}.
Parents with finite resources face difficult choices, and many appear to
balance quantity and quality in a way that confirms the trade-off
hypothesis.

Despite these similarities, there are some noteworthy nuances and gaps
in the ASEAN literature. One is the role of public policy and
development level. Evidence suggests that the Q--Q trade-off may be
mitigated in contexts with strong public support for education and
health. For instance, studies in developed countries (e.g.~Israel,
Norway) often find little or no trade-off once factors like birth order
are accounted for
\citep{black2005more, kristensen2010educational, angrist2010multiple}.
In Southeast Asia, however, public education quality and social safety
nets are still developing, and the cost of raising children (education
fees, food, etc.) is largely borne by families themselves
\citep{oecd2024sigi}. This may explain why the trade-off emerges so
clearly in the Philippines, Vietnam, Indonesia, and Thailand.

Another nuance is methodological: more recent studies employ credible
identification strategies (IVs, twins, policy shocks) and consistently
find a causal negative effect of family size on child outcomes, whereas
older studies without such controls sometimes found weaker effects or
none at all. This highlights the importance of accounting for
endogeneity.

Finally, there remain gaps for future research. Most ASEAN studies focus
on education and early-life health indicators as measures of child
quality; there is relatively little evidence on long-term outcomes such
as children's eventual earnings or income in adulthood. It is not yet
fully clear whether the schooling and health disadvantages observed in
larger families translate into significantly lower adult productivity or
income---a link that \citet{becker1976child} theorized but which could
be explored further in this region.

Additionally, while the trade-off appears pervasive, its magnitude can
vary: for example, the penalty of an extra child may be larger in poorer
rural areas than in urban or wealthier settings, suggesting that local
context (poverty, gender norms, access to services) can modulate the
trade-off. Comparative studies across ASEAN are still somewhat limited,
and a common challenge is disentangling related factors like birth
order, sibling composition, and parental preferences.

Nonetheless, the prevailing evidence from the Philippines and its
regional neighbors strongly supports Becker's quantity--quality
conjecture. As families have fewer children, they appear to invest more
in each child's education---investments crucial for human capital
development and economic growth. Conversely, high-fertility households
risk under-investing per child, which reinforces cycles of poverty and
inequality. This literature thus provides an important empirical
foundation for policies in the Philippines and ASEAN---from family
planning programs to education subsidies---that aim to ease the
quantity--quality trade-off and help families achieve both manageable
size and better outcomes for the next generation.
